A recipe for extremely reproducible enrichment analysis (and why we need it)

### 2023-06-22 manuscript/manuscript.Rmd
update article

Unreliable and irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles describing enrichment analysis were reproducible.
We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
reproducibility.
The findings indicate that current enrichment analysis practices are inadequate.
To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
a new project's enrichment analysis needs.
This recipe is built according to a framework we call "five pillars", which when implemented together
facilitates extreme reproducibility.

@@

In addition to shared code, data and containerisation, there are a few ways researchers can make their
work more reproducible.
Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Secondly, code and data can be linked. This means that the code knows the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one command
and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five
pillars (**Figure 2**).
Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
over a decade.

![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin. Republished with permission @Akalin2018-eg.](img/repro.webp "The computational research reproducibility spectrum.")

```{r,5pillars,fig.cap="Figure 2. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}

library("RColorBrewer")
cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
polygon(c(-0.2,5,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)

labs <- c("Code version\ncontrol &\npersistent\n sharing",
  "Compute \nenvironment \ncontrol",
  "Literate\nprogramming",
  "Documentation",
  "FAIR &\npersistent\ndata sharing")

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.3 , lab="End-to-end code coverage")
text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)

In this article, we focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[@Slonim2002-sf; @Khatri2012-do].
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertent misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script-based methods for this type
of work @Wijesooriya2022-li.
Although web-based tools are convenient, they are a potential reproducibility concern.
Firstly, incomplete reporting of methods, options and parameters is common.
Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available, hampering later reproducibility.
In some cases, the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay @Hennessey2013-tj.

@@

After conducting this reproduction study according to the methods described in the respective articles,
the 20 articles were classified based on the degree of reproducibility (Figure 3).
Six studies were classified as low reproducibility, while 10 were classified as medium and just three
were classified as high.
Justifications for these classifications are provided in Table 1.

```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 3. Examining reproducibility in a sample of 20 journal articles."}

@@

We have developed a step-by-step protocol for how to conduct an extremely reproducible
enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
protocols.io with a permissive license.
This protocol has the features of reproducibility mentioned in **Figure 1** and **Figure 2** including
version control with git and GitHub, compute environment control using Docker, literate programming
with Rmarkdown, persistent code and data sharing with Zenodo and documentation of all of these to
empower others to reproduce the workflow.

@@

The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
Rmarkdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
@Grolemund2017-zl p423.
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
@Zhao2023-ok.

```{r,diagram,echo=FALSE,fig.cap = "Figure 4. Flow chart of example workflow."}

@@

For quality control the number of reads per sample was calculated, and found to be in the range of
11-16M.
Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
genes.
Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
(Figure 5A).
Up and down-regulated gene sets were subjected to ORA separately, using the list of
13,168 genes as the background, and gene sets from REACTOME.

With ORA, we observed 149 up-regulated and 48 down-regulated gene sets.
We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo small expression changes.
This led to the identification of up-regulated pathways related to nuclear RNA export,
glucokinase regulation and G1/S DNA replication pathways; while down-regulated pathways
were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
(host defence) (Figure 5B).

After FCS, we identified 241 up-regulated gene sets and 155 down-regulated gene sets.
FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
and defective homologous recombination repair (Figure 5C).
Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.

Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
index of only 0.27 (Figure 5D).

![Figure 5. Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")

@@

We propose the five pillars (**Figure 2**), as a framework to guarantee computational reproducibility.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and Rmarkdown scripts to suit their
project's needs.
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we argue the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.

>>>>

Unreliable and irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles describing enrichment analysis were reproducible.
We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
reproducibility.
The findings indicate that current enrichment analysis practices are inadequate.
To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
a new project's enrichment analysis needs.
This recipe is built according to the "five pillars" framework, which facilitates extreme
reproducibility.

@@

In addition to shared code, data and containerisation, there are a few ways researchers can make their
work more reproducible.
Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Secondly, code and data can be linked. This means that the code knows the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one command
and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum @Akalin2018-eg, and the *five pillars*
framework has been described as a way of achieving extreme reproducibility @Ziemann2023-pt.
Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
over a decade.

In this article, we focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[@Slonim2002-sf; @Khatri2012-do].
While enrichment analysis is undoubtedly a key method for the interpretation of large omics
data sets, there are concerns that inadvertent misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script-based methods for this type
of work @Wijesooriya2022-li.
Although web-based tools are convenient, they are a potential reproducibility concern.
Firstly, incomplete reporting of methods, options and parameters is common.
Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available, hampering later reproducibility.
In some cases, the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay @Hennessey2013-tj.

@@

After conducting this reproduction study according to the methods described in the respective articles,
the 20 articles were classified based on the degree of reproducibility (Figure 1).
Six studies were classified as low reproducibility, while 10 were classified as medium and just three
were classified as high.
Justifications for these classifications are provided in Table 1.

```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility in a sample of 20 journal articles."}

@@

We have developed a step-by-step protocol for how to conduct an extremely reproducible
enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
protocols.io with a permissive license.
This protocol is in accordance with the *five pillars* including
version control with git and GitHub, compute environment control using Docker, literate programming
with Rmarkdown, persistent code and data sharing with Zenodo and documentation of all of these to
empower others to reproduce the workflow.

@@

The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
Rmarkdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
@Grolemund2017-zl p423.
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 2**).
For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
@Zhao2023-ok.

```{r,diagram,echo=FALSE,fig.cap = "Figure 2. Flow chart of example workflow."}

@@

For quality control the number of reads per sample was calculated, and found to be in the range of
11-16M.
Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
genes.
Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
(Figure 3A).
Up and down-regulated gene sets were subjected to ORA separately, using the list of
13,168 genes as the background, and gene sets from REACTOME.

With ORA, we observed 149 up-regulated and 48 down-regulated gene sets.
We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo small expression changes.
This led to the identification of up-regulated pathways related to nuclear RNA export,
glucokinase regulation and G1/S DNA replication pathways; while down-regulated pathways
were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
(host defence) (Figure 3B).

After FCS, we identified 241 up-regulated gene sets and 155 down-regulated gene sets.
FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
and defective homologous recombination repair (Figure 3C).
Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.

Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
index of only 0.27 (Figure 3D).

![Figure 3. Differential expression and enrichment analysis with the example workflow.](../img/fig_example.png "Example enrichment analysis on")

@@

This article is an attempt to foster increased reproducibility in genomics by demonstrating the
implementation of the five pillars principles into a commonly used routine, namely differential
expression and pathway enrichment analysis.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and Rmarkdown scripts to suit their
project's needs.
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we argue the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.
